---
title: "Chapter 3 - Data Transformation - Row"
output:
  pdf_document: default
  html_document: default
date: "2026-03-19"
---

```{r}
library(nycflights13)
library(tidyverse)
```

# Review the data

### 1. glimpse(flights)

### 2. view(flights)

### 3. print(flights)

# dplyr Basics

### 1. pip \|\> or %\>%

```{r warning=FALSE}
flights %>% 
  filter(dest == 'IAH') %>% 
  group_by(year, month, day) %>% 
  summarize(
    arr_delya = mean(arr_delay, na.rm = TRUE)
  )
```

# Rows - Most Important verbs:

### 1. filters(): keeps row based on the values of the columns

-   =; !=; \>; \>=;\<;\<=

-   %in% c(): check if in a list

### 2. arrange(): changes the order of the rows based on the value of the columns

-   desc()

### 3. distinct(): finds all the unique rows in a database.

-   .keep_all = TRUE

### 4. count()

-   sort = TRUE

### 5. head(10)

-   only show 10 rows

### 6. nrow(): show how many rows available

flights %\>%

nrow()

### 7. slice_max(distance, n = 5)

distinct(distance) %\>% \# select distinct first

slice_min(distance, n = 5) \# get the first 5 mins

### 8. select()

-   list of the columns

### 1. Filters()

-   =; !=; \>; \>=;\<;\<=

-   %in% c(): check if in a list

```{r}
# filter flight depart delay more than 120
# <, <=, >, >=, !=
flights %>% 
  filter(dep_delay > 120)
```

```{r}
# &, |
flights %>% 
  filter(month == 1 & day == 1)
```

```{r}
# %in%: keeps rows where the variable equals one of the vlaues on the right.
flights %>% 
  filter(month %in% c(1, 2))
```

```{r}
# asign to a new variable
jan1 <- flights %>% 
  filter(month == 1 & day == 1)
```

### 2. arrange()

-   desc(): sort by columns

```{r}
# desc(): decrease order
flights %>% 
  arrange(year, month, day, desc(dep_delay))
```

### 3. distinct()

-   .keep_all = TRUR: keep the other columns when filtering for unique rows

```{r}
flights %>% 
  distinct(month)

flights %>% 
  distinct(origin, dest)

flights %>% 
  distinct(origin, dest, .keep_all = TRUE)
```

### 4. count()

-   sort = TRUE

```{r}
flights %>% 
  count(origin, dest, sort = TRUE)
```

## Exercises

```{r}
names(flights)
```

-   Had an arrival delay of two or more hours

```{r}
flights %>% 
  filter(arr_delay > 120)
```

-   Flew to Houston (`IAH` or `HOU`)

```{r}
flights %>% 
  filter(dest %in% c('IAH','HOU'))
```

-   Were operated by United, American, or Delta

```{r}
flights %>% 
  filter(carrier %in% c('UA', 'AA', 'DA'))
```

```{r}
flights %>% 
  filter(month %in% c(7, 8, 9))
```

```{r}
flights %>% 
  filter(arr_delay >= 120 & dep_delay <= 0)
```

```{r}
flights %>% 
  filter(dep_delay >= 50 & dep_delay - arr_delay <= 30)
```

1.  **Sort `flights` to find the flights with the longest departure delays. Find the flights that left earliest in the morning.**

```{r}
names(flights)

```

```{r}
flights %>% 
  arrange(desc(dep_delay))

```

```{r}
flights %>% 
  arrange(dep_time) %>% 
  select(year, month, day, carrier, flight, dep_delay) %>% 
  head(10)
  

```

1.  **Sort `flights` to find the fastest flights. (Hint: Try including a math calculation inside of your function.)**

```{r}
flights %>% 
  mutate(miles_per_hour = distance / (air_time/60)) %>% 
  select(year, month, day, carrier, flight, distance, air_time, miles_per_hour) %>% 
  arrange(desc(miles_per_hour))

```

1.  **Was there a flight on every day of 2013?**

```{r}
flights %>% 
  distinct(year, month, day) %>% 
  nrow()

```

1.  **Which flights traveled the farthest distance? Which traveled the least distance?**

```{r}
flights %>% 
  arrange(desc(distance)) %>% 
  head(1)
```

```{r}
flights %>% 
  distinct(distance) %>% 
  slice_max(distance, n = 5)

```

```{r}
flights %>% 
  distinct(distance) %>% 
  slice_min(distance, n=5)
```

1.  Does it matter what order you used [`filter()`](https://dplyr.tidyverse.org/reference/filter.html) and [`arrange()`](https://dplyr.tidyverse.org/reference/arrange.html) if you’re using both? Why/why not? Think about the results and how much work the functions would have to do.

    -   the result would be the same

    -   the filter() -\> arrange() would be more efficient than arrange() -\> filter(). filter() -\> arrange() doesn't need to the arrange() the whole data set, only the ones after the filter()

```{r}

```

```{r}

```

```{r}

```
