---
title: "3 - Data Transformation - Columns"
output: html_document
date: "2026-03-20"
---

```{r}
library(tidyverse)
library(nycflights13)
```

## Columns

### 1. mutate()

-   Creates new columns that are derived from the existing columns

-   .before = 1: argument to instead add the variables to the left side, 1 means the first column.

-   .before = day: before the column day - day is a variable

-   .keep = 'used': only keep the used columns

### 2. select()

-   changes which columns are present

-   select(year, month, day): by columns

-   select(year:day): from year to day

-   select(! year:day): select all columns except year:day

-   select(where(is.character)): column data type is string

-   select(starts_with('year')): column name start with 'year'

-   select(ends_with('year')): column name end with 'year'

-   select(contains('year')): column name contain 'year'

-   select(num_range('x',1:3)): matches: x1, x2, and x3

-   select(tail_num = tailnum): rename column name from 'tailnum' to 'tail_num', this would only show the column 'tail_nam'

### 3. rename()

-   changes the names of columns, and show all columns with the renamed one

-   rename(tail_num = tailnum)

### 4. relocate()

-   changes the positions of the columns

-   relocate(year): will move column 'year' to the first column

-   relocate(year, .before=day): move year to before day

-   relocate(starts_with('arr'), .before=dep_time): move columns starts with 'arr' before 'dep_time'

### 1. mutate()

```{r}
flights %>% 
  mutate(
    gain = dep_delay - arr_delay, # gain by the speed
    speed = distance / air_time * 60,
    .before = 1
  )
```

```{r}
flights %>% 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60, 
    gian_per_hour = gain / hours,
    .keep='used'
  )
```

### 2. Select()

```{r}
flights %>% 
  select(year, month, day) %>% 
  head(5)
```

```{r}
flights %>% 
  select(year:day) %>% 
  head(5)
```

```{r}
flights %>% 
  select(! year:day) %>% 
  head(5)
```

```{r}
flights %>% 
  select(where(is.character))
```

```{r}
flights %>% 
  select(starts_with('year'))
```

```{r}
flights %>% 
  select(tail_num = tailnum)
```

```{r}
flights %>% 
  rename(tail_num = tailnum)
```

```{r}
flights %>% 
  relocate(time_hour, air_time)
```

```{r}
flights %>% 
  relocate(year:dep_time, .after = time_hour)
```

```{r}
flights %>% 
  relocate(starts_with('arr'), .before = dep_time)
```

### Exercises

1.  **Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?**

```{r}
flights %>% 
  mutate(delay = 
           (trunc(dep_time/100)  * 60 + (dep_time/100 - trunc(dep_time/100))*100) -
           (trunc(sched_dep_time/100)  * 60 + (sched_dep_time/100 - trunc(sched_dep_time/100))*100)
         ) %>% 
  select(dep_time, sched_dep_time, delay, dep_delay)
```

1.  **Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay`from `flights`.**

```{r}
flights %>% 
  head(5)
```

```{r}
flights %>% 
  select(dep_time, dep_delay, arr_time, arr_delay) %>% 
  head(5)
```

```{r}
flights %>% 
  select((contains('dep') | contains('arr')) & !contains('sched') & !contains('carr')) 
```

```{r}
```

```{r}
```
