---
title: "3 - Data Transformation - Columns"
output: html_document
date: "2026-03-20"
---

```{r}
library(tidyverse)
library(nycflights13)
```

## Part 1 - Columns

### 1. mutate()

-   Creates new columns that are derived from the existing columns

-   .before = 1: argument to instead add the variables to the left side, 1 means the first column.

-   .before = day: before the column day - day is a variable

-   .keep = 'used': only keep the used columns

### 2. select()

-   changes which columns are present

-   select(year, month, day): by columns

-   select(year:day): from year to day

-   select(! year:day): select all columns except year:day

-   select(where(is.character)): column data type is string

-   select(starts_with('year')): column name start with 'year'

-   select(ends_with('year')): column name end with 'year'

-   select(contains('year')): column name contain 'year';

-   select(contains('year', ignore.case=FALSE): default case-insensitive

-   select(num_range('x',1:3)): matches: x1, x2, and x3

-   select(tail_num = tailnum): rename column name from 'tailnum' to 'tail_num', this would only show the column 'tail_nam'

### 3. rename()

-   changes the names of columns, and show all columns with the renamed one

-   rename(tail_num = tailnum)

### 4. relocate()

-   changes the positions of the columns

-   relocate(year): will move column 'year' to the first column

-   relocate(year, .before=day): move year to before day

-   relocate(starts_with('arr'), .before=dep_time): move columns starts with 'arr' before 'dep_time'

## Part 2 - The pip

## Part 3 - group_by()

-   use group_by() to divide your dataset into groups meaningful for your analysis

### 1. summarize()

-   after the group_by()

-   summarize(

    -   delay = mean(dep_delay, na.rm=TRUE),

    -   n = n() \# show total rows for each group

    -   )

### 2. The slice_Functions

-   slice_head(n=1): select top 1; slice_head(prop=0.1): select top 10%

-   slice_tail(n=1)

-   slice_max(x, n=1, ***with_ties***=FALSE): with_ties = FALSE - if two have the same max, only keep one row, keeps which ever appears first in the data.

-   slice_min(x, n=1)

-   slice_sample(n=1)

### 3. Grouping by Multiple Variables

### 4. ungroup()

### 5. .by():

-   Experimental syntax for per-operation grouping

-   .by works with all verbs and has the advantage that you don't need to use the .groups argument to suppress the grouping message or ungroup() when you're done.

### 1. mutate()

```{r}
flights %>% 
  mutate(
    gain = dep_delay - arr_delay, # gain by the speed
    speed = distance / air_time * 60,
    .before = 1
  )
```

```{r}
flights %>% 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60, 
    gian_per_hour = gain / hours,
    .keep='used'
  )
```

### 2. Select()

```{r}
flights %>% 
  select(year, month, day) %>% 
  head(5)
```

```{r}
flights %>% 
  select(year:day) %>% 
  head(5)
```

```{r}
flights %>% 
  select(! year:day) %>% 
  head(5)
```

```{r}
flights %>% 
  select(where(is.character))
```

```{r}
flights %>% 
  select(starts_with('year'))
```

```{r}
flights %>% 
  select(tail_num = tailnum)
```

```{r}
flights %>% 
  rename(tail_num = tailnum)
```

```{r}
flights %>% 
  relocate(time_hour, air_time)
```

```{r}
flights %>% 
  relocate(year:dep_time, .after = time_hour)
```

```{r}
flights %>% 
  relocate(starts_with('arr'), .before = dep_time)
```

### Exercises

1.  **Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?**

```{r}
flights %>% 
  mutate(delay = 
           (trunc(dep_time/100)  * 60 + (dep_time/100 - trunc(dep_time/100))*100) -
           (trunc(sched_dep_time/100)  * 60 + (sched_dep_time/100 - trunc(sched_dep_time/100))*100)
         ) %>% 
  select(dep_time, sched_dep_time, delay, dep_delay)
```

1.  **Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay`from `flights`.**

```{r}
flights %>% 
  head(5)
```

```{r}
flights %>% 
  select(dep_time, dep_delay, arr_time, arr_delay) %>% 
  head(5)
```

```{r}
flights %>% 
  select((contains('dep') | contains('arr')) & !contains('sched') & !contains('carr')) 
```

```{r}
select(flights, any_of(c('dep_time','dep_delay','arr_time','arr_delay','flights')))
```

1.  **What happens if you specify the name of the same variable multiple times in a [`select()`](https://dplyr.tidyverse.org/reference/select.html) call?**

```{r}
flights %>% 
  select('year','year') %>% 
  print(width=Inf)
```

What does the [`any_of()`](https://tidyselect.r-lib.org/reference/all_of.html) function do? Why might it be helpful in conjunction with this vector?

```         
variables <- c("year", "month", "day", "dep_delay", "arr_delay")
```

```{r}
variables <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, any_of(variables)) %>% 
  print(width=Inf)
```

Does the result of running the following code surprise you? How do the select helpers deal with upper- and lowercase by default? How can you change that default?

```{r}
flights |> select(contains("TIME", ignore.case=TRUE)) %>% 
  print(width=50)
```

1.  **Rename `air_time` to `air_time_min` to indicate units of measurement and move it to the beginning of the data frame.**

```{r}
flights %>% 
  rename(air_time_min = air_time) %>% 
  relocate(air_time_min, .before=1) %>% 
  print(width=100)
```

## The Pipe

find the fast flights to Houston’s IAH airport: you need to combine [`filter()`](https://dplyr.tidyverse.org/reference/filter.html), [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html), [`select()`](https://dplyr.tidyverse.org/reference/select.html), and [`arrange()`](https://dplyr.tidyverse.org/reference/arrange.html):

```{r}
flights %>% 
  filter(dest == 'IAH') %>% 
  mutate(speed = distance / air_time * 60) %>% 
  select(year:day, dep_time, carrier, flight, speed) %>% 
  arrange(desc(speed)) %>% 
  print(width=Inf)
```

## group_by()

```{r}
flights %>% 
  group_by(month) %>% 
  print(width=50)
```

## summarize()

```{r}
flights %>% 
  group_by(month) %>% 
  summarize(
    avg_delay = mean(dep_delay, na.rm=TRUE)
  ) %>% 
  print(width=Inf)
```

```{r}
flights %>% 
  group_by(month) %>% 
  summarize(
    delay = mean(dep_delay, na.rm=TRUE),
    n = n()
  ) %>% 
  print(width = Inf)
```

## The slice_Functions

```{r}
flights %>% 
  group_by(dest) %>% 
  slice_max(arr_delay, n=1, with_ties=FALSE) %>% 
  relocate(dest, arr_delay) %>% 
  print(width=60)
```

## Grouping by multiple variables

```{r}
daily <- flights %>% 
  group_by(year, month, day)
daily %>% 
  print(width=60)
```

```{r}
daily %>% 
  summarize(n = n(),
            .groups = 'drop_last') %>% 
  print(width = 50)
```

## Ungrouping

```{r}
daily %>% 
  ungroup()
```

## .by

```{r}
flights %>% 
  summarize(
    delay = mean(dep_delay, na.rm=TRUE),
    n = n(),
    .by = month
  )
```

```{r}
flights %>% 
  summarize(
    delay = mean(dep_delay, na.rm=TRUE),
    n = n(),
    .by = c(origin, dest)
  )
```

## Exercises

```{r}
flights
```

```{r}
flights %>% 
  group_by(carrier) %>% 
  summarize(
    delay = mean(dep_delay, na.rm=TRUE)
  ) %>% 
  slice_max(delay)
```

1.  **Find the flights that are most delayed upon departure from each destination.**

```{r}
flights %>% 
  group_by(dest) %>% 
  slice_max(dep_delay, n = 1, with_ties=FALSE) %>% 
  select(dest, carrier, flight, dep_delay) %>% 
  arrange(desc(dep_delay))
```

1.  **How do delays vary over the course of the day. Illustrate your answer with a plot.**

```{r}
flights %>% 
  group_by(hour) %>% 
  summarize(
    delay = mean(dep_delay, na.rm=TRUE)
  ) %>% 
  ggplot(aes(x = hour, y = delay)) +
  geom_point()+
  geom_smooth()+
  labs(
    title = 'Average Departure Delay by Hour of Day',
    x = 'Hour of Day',
    y = 'Average Departure Delay (mins)'
  )

```

1.  **Explain what [`count()`](https://dplyr.tidyverse.org/reference/count.html) does in terms of the dplyr verbs you just learned. What does the `sort` argument to [`count()`](https://dplyr.tidyverse.org/reference/count.html) do?**

```{r}
flights %>% 
  count(year, month, sort = TRUE) %>% 
  print(width=Inf)
```

```{r}
df <- tibble(
  x = 1:5,
  y = c("a", "b", "a", "a", "b"),
  z = c("K", "K", "L", "L", "K")
)
```

```{r}
df
```

```{r}
df %>% 
  group_by(y)
```

```{r}
df %>% 
  arrange(y)
```

```{r}
df %>% 
  group_by(y) %>% 
  summarize(mean_x = mean(x))
```

```{r}
df %>% 
  group_by(y, z) %>% 
  summarize(mean_x = mean(x))
```

```{r}
df %>% 
  group_by(y, z) %>% 
  summarize(mean_x = mean(x), .groups = 'drop')
```

```{r}
df %>% 
  group_by(y, z) %>% 
  summarize(mean_x = mean(x), .groups = 'drop')
```

```{r}
df %>% 
  group_by(y, z) %>% 
  summarize(mean_x = mean(x))
```

```{r}
df %>% 
  group_by(y, z) %>% 
  mutate(mean_x = mean(x))
```

## Case Study: Aggregates and Sample Size

```{r}
batters <- Lahman::Batting |> 
  group_by(playerID) |> 
  summarize(
    performance = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    n = sum(AB, na.rm = TRUE)
  )
batters
```

```{r}
batters %>% 
  filter(n > 100) %>% 
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 1/ 10) +
  geom_smooth(se = FALSE)
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
